Introduction Concurrent Programming with CXL

### 2020-05-15 book/paper.tex
Latest

@@

A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable list of CXL machine instructions.  The
instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable tag, a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL.
\end{itemize}

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
By convention, both the memory and the register of a process always contain
a special type of CXL value in which the root is a dictionary
that maps strings to CXL values.  It represents the state of a collection of
variables named by the strings.

>>>>

\begin{figure}
\begin{code}
\begin{verbatim}
Count.cxl:1 def up():
    0 Jump 22
    1 Frame ('up', 'Count.cxl', 1, 5) None 10
Count.cxl:2     count = count + 1;
    2 Constant (1, 'Count.cxl', 2, 21)
    3 Load ('count', 'Count.cxl', 2, 13)
    4 Nary ('+', 'Count.cxl', 2, 19) 2
    5 Name ('count', 'Count.cxl', 2, 5)
    6 Store 1
Count.cxl:3     upDone = True;
    7 Constant (True, 'Count.cxl', 3, 14)
    8 Name ('upDone', 'Count.cxl', 3, 5)
    9 Store 1
   10 Return
   11 ...
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:updown}.}
\label{fig:updowncode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
Assertion failed ('assert', 'Count.cxl', 17, 1) 1

==== Safety violation ====
__main__ 0 [0,22-36]     37 { 'count':  0, 'upDone': False, 'downDone': False }
up       0 [1-5]          6 { 'count':  0, 'upDone': False, 'downDone': False }
down     0 [12-20]       21 { 'count': -1, 'upDone': False, 'downDone': True  }
up       0 [6-9]         10 { 'count':  1, 'upDone': True,  'downDone': True  }
__main__ 0 [37-41,43-48] 48 { 'count':  1, 'upDone': True,  'downDone': True  }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:updown}.}
\label{fig:updownoutput}
\end{figure}

@@

A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable tag, a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL.
\end{itemize}

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
By convention, both the memory and the register of a process always contain
a special type of CXL value in which the root is a dictionary
that maps strings to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the strings.

Directories can be organized into a tree.  Each node in the tree is then identified
by a sequence of strings, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value.

Compiling the code in Figure~\ref{fig:updown} results in the CXL machine code
listed in Figure~\ref{fig:updowncode}.
The CXL machine is predominantly a \emph{stack machine}.
Most instructions pop values from the stack or push values onto the stack.
Initially there is one process that starts executing at instruction 0.
In this case, there is a \texttt{JUMP} instruction that sets the program
counter to 22.  At program counter 1 is the code for the \texttt{up} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item A token containing the name of the method;
\item A token containing the argument of the method (in this case the
method has no named argument);
\item A program counter of the method's \texttt{Return} instruction.
\end{enumerate}

A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
line number in the file, and a column number in the file.  For example,
\texttt{(’up’, ’Count.cxl’, 1, 5)} represents the lexeme \texttt{up} in
file \texttt{Count.xcl}, line 1, column 5.  Tokens make it easy to relate
the CXL machine instructions to the CXL source code.

The code generated from \texttt{count := count + 1} in line 2 of
\texttt{Count.cxl} is as follows:

\begin{enumerate} \setcounter{enumi}{1}
\item the \texttt{Constant} instruction pushes the constant 1
onto the stack of the process.
\item The \texttt{Load} instruction loads the shared variable
\texttt{count} from the shared memory and pushes its value onto the
stack.
\item \texttt{Nary} stands for \texttt{n-ary}: in this case it is a \texttt{+}
operation with 2 arguments.  It pops two values from the stack, adds them,
and pushes the result back onto the stack.
\item The \texttt{Name} instruction pushes the address of the \texttt{count}
variable onto the stack.
\item The \texttt{Store} instruction pops the address of a variable and pops
a CXL value, and updates the memory by replacing the CXL value oat the given
address with the new CXL value.
\end{enumerate}

Figure~\ref{fig:updownoutput} shows the output produced by running the
\texttt{Count.cxl} program.
It starts by reporting that the assertion on line 17 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Count.cxl’, 17, 1) 1}.
Next it reports an execution that failed this assertion.  The output has
five columns:
\begin{enumerate}
\item The entry point of the process (\texttt{\_\_main\_\_} for the main process);
\item The tag of the process (in this case always 0);
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

If we look at the middle three rows, we see that:
\begin{enumerate}
\item \texttt{up} executed instructions 1 through 5, loading the value of
\texttt{count} but stopping just before storing the value 1 into \texttt{count};
\item \texttt{down} executed instructions 12 through 20, which is all of the
\texttt{down} method, loading the value 0 from \texttt{count} and storing $-1$
into \texttt{count};
\item \texttt{up} continues execution, storing the value 1 into \texttt{count}.
\end{enumerate}

This makes precise the concurrency problem that we encountered.
